---
title: "Professor Evaluations"
author: "William Duquette"
date: "Due on Wednesday, April 22 at 11:59PM"
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
---

```{r setup, echo=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


```{r,echo=FALSE, include=FALSE}
require(car)
require(carData)
require(ggplot2)
require(tidyverse)
require(EnvStats)
require(MASS)
library(MuMIn)
require(gridExtra)
require(Hmisc)
library(expss)
```

```{r,echo=FALSE, include=FALSE}
download.file("http://www.openintro.org/stat/data/evals.RData", destfile = "evals.RData")
load("evals.RData")

drops = c("bty_f1lower", "bty_f1upper", "bty_f2upper", "bty_m1upper", "bty_m2upper", "bty_m1lower")
evals = evals[, !(names(evals) %in% drops)]

evals2=evals
evals2$cls_did_eval=log(evals$cls_did_eval)
evals2$cls_students=log(evals$cls_students)

evals=apply_labels(evals, cls_did_eval="Number of Students who Completed Evaluations", cls_students= "Total Number of Students in Class")
evals2=apply_labels(evals2, cls_did_eval="Number of Students who Completed Evaluations", cls_students= "Total Number of Students in Class")
```

Project Description: It is very important for schools to be able to receive accurate and fair evaluations of professors. If their are factors outside of a professors teaching skills that are affecting a professors score, then it is important to acknowledge them to avoid the termination of a professor who got a negative evaluation because of something they cannot control, such as ethnicity or gender. A study conducted at the University of Austin pooled evaluations from 463 professors at their campus, and asked six students to rate the professors on their physical appearance. The main question that is being asked is: What non-teaching characteristics affect a professors evaluation score. I will create a model that predicts, which of the 15 variables, are important to estimate the factors and their significance on a professors score.


Relevant Variables: There are 15 variables that are inside the data set, but not all of them are significant. For this question, score will be the response variable. This variable, score, measures the average professor evaluation score from 1 to 5 (1 being very unsatisfactory and 5 being excellent) (see below). The explanatory variables that will be included in the model are listed below:

1. Rank - This categorical variable denotes whether the professor is teaching, on the tenure track, or tenured. Below is a graph that shows the distribution of the explanatory variable. As you can see there are more tenured professors than any other rank, while the split between teaching and tenure track is about the same. 

```{r}
ggplot(evals, aes(x=rank))+geom_bar()+labs(title = "Distribution of Rank")
summary(evals$rank)
```


2. Ethnicity - This categorical variable denotes whether the professor is a minority or not. Below is a graph that shows the distribution of the explanotory variable. As you can see there are significantly more not minority professors than minority professors.

```{r}
ggplot(evals, aes(x=ethnicity))+geom_bar()+labs(title="Distribution of Ethnicity")
summary(evals$ethnicity)
```


3. Gender - This categorical variable denotes whether the professor is male or female. Below is a graph that shows the distribution of the explanotory variable. As you can see there are 73 more male professors than female professors. 

```{r}
ggplot(evals, aes(x=gender))+geom_bar()+labs(title="Distribution of Gender")
summary(evals$gender)
```


4. Language - This categorical variable denotes whether the school that the professor recieved their education at taught classes in English or not. Below is a graph that shows the distribution of the explanotory variable. As you can see there are significantly more professors that took classes in English than professors that took classes not taught in English.

```{r}
ggplot(evals, aes(x=language))+geom_bar()+labs(title="Distribution of Language")
summary(evals$language)
```


5. Age - This variable measures the age of the professor. Below is a graph that shows the distribution of the explanotory variable. As you can see the distribution of age is roughly normal. 

```{r}
ggplot(evals2, aes(x=age))+geom_histogram(bins=10)+labs(title="Distribtuion of Age")
```


6. cls_perc_eval - This variable measures the percent of students in a class that completed an evaluation. Below is a graph that shows the distribution of the explanatory variable. The distribution on this variable is slightly skewed left but not enough to warrant a transformation. Given that more students fill out a survey as opposed to not fill out a survey, this distribution makes sense. 

```{r}
ggplot(evals, aes(x=cls_perc_eval))+geom_histogram(bins=10)+labs(title="Distribution of cls_perc_eval", x="Percent of Students that Completed an Evaluation")
```

7. cls_did_eval - This variables meaures the amount of students that completed the evaluation. Below is a graph that shows the distribution of the explanotory variable. I have applied the log() transformation to the cls_did_eval variable. See below.

```{r,message=FALSE}
ggplot(evals, aes(x=cls_did_eval))+geom_histogram(bins=10)+labs(title="Distribution of cls_did_eval", x="Amount of Students that Completed an Evaluation")
```


8. cls_students - This variable measures the amount of students in the class. Below is a graph that shows the distribution of the explanotory variable. I have applied the log() transformation to the cls_students variable. See below.

```{r,message=FALSE}
ggplot(evals, aes(x=cls_students))+geom_histogram(bins=10)+labs(title="Distribution of cls_students",x="Amount of Students in a Class")
```

9. cls_level - This categorical variable denotes whether a class level is lower or upper. Below is a graph that shows the distribution of the explanotory variable. As you can see there roughly 50% more upper level classes than lower level classes. 

```{r}
ggplot(evals, aes(x=cls_level))+geom_bar()+labs(title="Distribution of cls_level",x="Class Level")
summary(evals$cls_level)
```

10. cls_profs - This categorical variable denotes whether the number of professors teaching sections in a course in the sample is single or multiple. Below is a graph that shows the distribution of the explanotory variable. As you can see there are roughly 50% more multiple than single. 

```{r}
ggplot(evals, aes(x=cls_profs))+geom_bar()+labs(title="Distribution of cls_profs",x="Multiple or Single")
summary(evals$cls_profs)
```

11. cls_credits - This categorical variable denotes whether the course being taught was one credit or multi-credit courses. Below is a graph that shows the distribution of the explanotory variable. As you can see there are significantly more multi-credit courses. 

```{r}
ggplot(evals, aes(x=cls_credits))+geom_bar()+labs(title="Distribution of cls_credits",x="Whether the Course is One Credit or Multi-Credit")
summary(evals$cls_credits)
```


12. bty_avg - This variables measures the average beauty rating of a professor. Below is a graph that shows the distribution of the explanotory variable. The distribution of this variable is approximately normal.

```{r}
ggplot(evals2, aes(x=bty_avg))+geom_histogram(bins=10)+labs(title="Distribution of bty_avg",x="Professors Average Beauty Average")
```


13. pic_outfit - This categorical variable denotes whether the professor was wearing a formal or informal outfit for their picture. Below is a graph that shows the distribution of the explanotory variable. As you can see there are significantly more not formal outfits in pictures. 

```{r}
ggplot(evals, aes(x=pic_outfit))+geom_bar()+labs(title="Distribution of pic_outfit", x="Whether the Professor was Dressed Formally or Informally")
summary(evals$pic_outfit)
```


14. pic_color - This categorical variable denotes whether the professors picture was in color or black & white. Below is a graph that shows the distribution of the explanotory variable. As you can see there are significantly more color pictures of professors.

```{r}
ggplot(evals, aes(x=pic_color))+geom_bar()+labs(title="Distribution of pic_color", x="Whether the Professor's Pictures was in Color or Black & White")
summary(evals$pic_color)
```




Exploratory Data Analysis: The first graph shown is a histogram showing the distribution of the response variable: score. The data is slightly skewed left, but not enough to warrant a transformation. The second set of graphs, which uses the original data set, is a scatter plot matrix that shows each possible explanatory variable's relationship to the response variable. As you can see *cls_did_eval* and *cls_students* are skwewed right and need to be transformed. 

```{r,echo=FALSE}
# Analyzing all of variables to check if they need to be transformed
par(mfrow=c(2,2))
ggplot(evals2, aes(x=score))+geom_histogram(bins=15)+labs(title = "Distribution of response variable: score")

plot(score~.,data=evals,  main="Score vs. Explanatory Variable")
```

Below you can see the analysis that is done on *cls_did_eval* and *cls_students* to see what transformation is necessary. I have created graphs that the show the variables before and after they have been transformed. The transformation applied was log(). The reason that this transformation was used was because both variables cover several orders of magnitude. 

```{r,echo=FALSE,message=FALSE}
# Chekcing cls_did_eval

before1<-ggplot(evals, aes(x= cls_did_eval, y= score))+geom_point()+labs(title= "Before", x= "Number of Students who Completed Evaluations")

after1<-ggplot(evals2, aes(x=cls_did_eval, y= score))+geom_point()+labs(title="After", x="Number of Students who Completed Evaluations")

grid.arrange(before1,after1,ncol=2)

```

```{r, echo=FALSE, message=FALSE}
# Checking cls_students

before2<-ggplot(evals, aes(x=cls_students, y=score)) + geom_point()+labs(title= "Before Transformation for cls_students", x= "Total Number of Students in Class")

after2<-ggplot(evals2, aes(x=cls_students, y=score))+geom_point()+labs(title="After Transformation for cls_students", x="Total Number of Students in Class")

grid.arrange(before2,after2)
```



Results: I have found two interaction terms that are statistically significant. An interaction term is a variable that is a function of the current explanatory variables. An interaction term is appropiate inside a model if one explanatory variable is dependent on another explanatory variable. For my first interaction term, age*bty_avg, you can see that the older a professor is, and the higher the beauty average a professor has, there is a greater significance on their score. 

```{r}
# Forming my first interaction term

ggplot(evals2, aes(x=bty_avg, y=score, color=age))+geom_point()
mod.int.1<-lm(score~age*bty_avg, data=evals2)
summary(mod.int.1)
```

```{r, include=FALSE}
newdata= data.frame(bty_avg=c(4,4,4,4,6,6,6,6,8,8,8,8,10,10,10,10), age=c(40,50,60,70,40,50,60,70,40,50,60,70,40,50,60,70))
predict(mod.int.1,newdata)
```
The above summary shows the coefficients of the model that shows the interaction term between bty_avg and age. Given that both a professors beauty average and their age are numeric values, I will use a table to best illustrate the significance of the interaction variable. The horizontal columns show the beauty average, and the vertical columns show the professors age. As you can see from this table, it is clear that the more attractive a teacher is rated the more likely it is that a teachers score will increase as the professor's age increases.

                            Table 1
                           ---------------------------------------------------------------
                            |           |     4     |      6     |     8     |     10     |
                           ---------------------------------------------------------------
                            |   40      | 4.210570  | 4.260383   | 4.310197  | 4.360010   |
                           ---------------------------------------------------------------
                            |   50      | 4.161944  | 4.318161   | 4.474327  | 4.630494   |
                           ---------------------------------------------------------------
                            |   60      | 4.113418  | 4.375938   | 4.638458  | 4.900978   |
                           ---------------------------------------------------------------
                            |   70      | 4.064841  | 4.433715   | 4.802588  | 5.171462   |
                          ---------------------------------------------------------------

                          




For my second interaction term I found that there was an increased effect on a professors score if the professor was both a minority and tenured, on the tenure track, or teaching. As you can see from the graph minority professors are judged more harshly than their non-minority counterparts on the tenure and tenure track. One thing that is important to note is that teaching is the only category where professors who are a minority are scored higher. One reason that this might be the case is because of the relatively small sample size. The sample size for both minority and teaching professors is 10, compared to 28 for tenure track and minority and 26 for tenured and minority professors. 

```{r}
#Forming my second interaction term

ggplot(evals2, aes(x=rank, y= score, fill=ethnicity))+geom_boxplot()
mod.int.2<- lm(score~rank*ethnicity, data= evals2)
summary(mod.int.2)
table(evals2$ethnicity, evals2$rank)
```

The above summary shows the coefficients of a model that was made to analyze the significance of the interaction term between rank and ethnicity. This table shows what each coefficient in the above summary represent:

Table 2

Variable                                | Effect 
--------------------------------------  | ------------------------------------------------------------------------------------------------------------------
`ranktenure track` & `ranktenured`      | On average tenure track professors and tenured professors received a score of 0.9100 and 0.8369, respectively, less than their teaching counterparts.
`ethnicitynotminority`                  | On average non minority professors received a score of 0.5828 less then their minority counterparts.
`ranktenure track: ethnicitynotminority`| On average professors who were both on the tenure track and not a minority received a score of 0.9266 more than their counterparts that are a minority and not on the tenure track. If the professor was on a tenure track and not a minority the score would increase 0.9266 in addition to the affect of being on the tenure track (-0.9100) and not being a minority (-0.5828)
`ranktenured:ethnicitynot minority`     |  On average professors who were both tenured and not a minority received a score of 0.7679 more than their counterparts that are a minority and not tenured. If the professor was tenured and not a minority then there score would increase by 0.7679 in addition to the affect of being tenured (-0.8369) and not being a minority (-0.5828)


I used to backwards stepwise model selection to build my model, which is an algorithm that finds the best model for predicting the average evaluation score. Having the step function be in the backwards direction means that the model will start with all the variables and will remove variables as the algorithm sees fit. It compares the models that it produces using the AIC scores from each model; the lower the score the better. 

```{r}
mod.full<-lm(score~.+age*bty_avg+rank*ethnicity, data=evals2)
mod.back<-step(mod.full,direction="backward", trace=0)
summary(mod.back)
```

This is the model that was created from the stepwise algorithm. As you can see there are 10 explanatory variables. This model has an Adjusted R-squared value of 0.1807. An Adjusted R-squared compares the explanatory power of a models that contain varying amounts of explanatory variables. See more on the adjusted R-squared in the Discussion section. The estimate coloum shows how much score would increase or decrease if the other variables are held constant. Below is an explanation of each of the coefficient:


variable                                | description
----------------------------------------|-----------------------------------------------------------------------------------------
`ranktenure track`& `ranktenured`       | Refer to table 2
`ethnicitynot minority`                 | Refer to table 2
`gendermale`                            | On average male professors recieved a score of 0.194641 more than their female counterparts.
`languagenon-english`                   | On average non-english professors recieved a score of 0.226024 less then their english counterparts.
`age`                                   | Refer to table 1
`cls_perc_eval`                         | On average for every 1 percent increase in the amount of students who completed an evaluation for a professors class, their score would increase by 0.004604.
`cls_creditsone credit`                 | On average professors who taught one credit courses recieved a score of 0.343379 more than their counterparts that taught multicredit courses.
`bty_avg`                               | Refer to table 1
`pic_outfitnot formal`                  | On average professors with informal outfits in their pictures recieved a score of 0.122766 less than their counterparts that had formal outfits on in their picture.
`pic_colorcolor`                        | On average professors with color pictures revieved a score of 0.146570 less than their counterparts that had black and white pictures.
`age:bty_avg`                           | Refer to table 1
`ranktenure track:ethnicitynot minority`| Refer to table 2
`ranktenured:ethnicitynot minority`     | Refer to table 2




When you made a linear model there are certian assumptions that have to be made. The first assumption is that the data is linearly related; you can see from plots above that the data is linearly related. The second assumption is that the variables are independent, whichin this data set they are. The third assumption is that the residuals should be normally distributed. The Normal Q-Q plot on the top left corner shows that the data is normally distributed. The points fit along the dotted line for almost all of the line, only veering of the dotted line at the very ended. The final assumption is that there is equal variance of the residuals. The Residuals vs Fitted graph in the top left corner shows that the data has equal variance. The assumptions for a linear model are met. 

```{r}
plot(mod.back)
```

Discussion: Fortunately, many probelms that statistical studies face do not apply in this case. The data was complete and fairly cleaned. There were only two variables that had to be transformed. One problem that I would point out is that this data includes no variables that measure a professors teaching ability. Ideally, there would be more variables that could measure this. For example, there could be a variables that measures how well a professor explains concepts, how fairly a professor grades, and how available a professor is to meet a student, to name a few. I understand that this study was conducted to see what effect attractiveness had on a professors score, but if variables like these were present you would be able to tell how important such things are to a professors score. Or in other words, you would be able to see which group of variables had more weight on a professors score: attractiveness or teaching ability. What is alarming about this study is that aspects about yourself that a professor cannot control can significantly affect your score. For example, a tenured professor that was also not a minority would recieve a score of 0.9266 more than their minority and non-tenure track counterparts. Another example is the beuaty average. It is is clear that the more attractive you are, the higher score you receive (see table 1). One final example of this alarming trend is gender. If you are a male professor you would recieve a score of 0.194641 more than your female counterparts. Given that the adjusted R-squared is in the generally accepted range, it would suggest that attractiveness has a significant effect on score.

Conclusion: Given that this data does not include any variables that measure a professors teaching ability, I would reccomend that changes be made to how a professor is evaluated. It is unfair for a professor to be penalized for something that they cannot control such as ethnicity, gender, or ehtnicity. It is foolish to judge a professor on their score if things that they cannot control affect it. Noting that, I would reccomend that this study be repeated, but this time include variables that measure teaching ability, such as the suggestions in the discussion category. This would allow someone to properly weigh which variables matter the most. Knowing that attrativeness effects someones score should be noted, and less significance should be placed on a professors score until such a time where a proffesor can be properly ranked on their teaching ability. 






